---
title: "SMT_Ch10"
author: "Daehyeog Lee"
format: html
editor: visual
---

```{r}
library(compute.es); library(car); library(ggplot2); library(multcomp);
library(pastecs); library(WRS2)
```

## SMT1

Factor: Punish, Reward, Indifferent

Variable: Percentage

Hypothesis 1: Reward results in better exam results than either punishment or indifference.

Hypothesis 2: Indifference will lead to significantly better exam results than punishment.

```{r}
teachData <- read.delim("Teach.dat", header = TRUE)
head(teachData)
teachData$group <- factor(
  teachData$group, levels = c(1:3), labels = c("Punish","Indifferent", "Reward")
  )

head(teachData)

teachDatascatter <- ggplot(teachData, aes(group, exam))
teachDatascatter +
  geom_point() +
  stat_summary(
    fun = mean, geom = "line", size = 1, color = "black"
    ) +
  geom_smooth(method = "lm", aes(), color = "Red") +
  stat_summary(
    fun.data = mean_cl_boot, geom = "errorbar",
    width = 0.2, size = 0.75, color = "blue"
    ) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "blue") +
  stat_summary(fun = mean, geom = "point", size = 3, color = "red") +
  labs(x = "Groups", y = "Percentage")

leveneTest(teachData$exam, teachData$group, center = median)

teachModel <- aov(exam ~ group, data = teachData)
summary(teachModel)
```

Levene's test was not significant, which indicates the assumption of homogeneity of variance has met.

The result of `summary(teachModel)` tells us that there was a significant effect of teaching style on exam marks. However, we don't know what the exact effect was yet.

```{r}
contrast1 <- c(1, 1, -2)
contrast2 <- c(1, -1, 0)
contrasts(teachData$group) <- cbind(contrast1, contrast2)
teachData$group
```

The first contrast compares `Reward` and against `Punish` and `Indifferent`. The second contrast compares `Punish` and `Indifferent`.

```{r}
teachPlanned <- aov(exam ~ group, data = teachData)
summary.lm(teachPlanned)
```

The output tells us that the exam mark from `Reward` was significantly different from that of `Punish` and `Indifferent` method. Also, the exam mark from `Punish` was significantly different from that of `Indifferent` method.

```{r}
rcontrast <- function(t, df)
  {
  r <- sqrt(t^2/(t^2+df))
  print(paste("r = ", r))
  }
omegaSquared <- function(ssm, ssr, msr, dfm)
  {
  omega <- (ssm - dfm*msr)/(ssm + ssr + msr)
  print(paste("omega^2 = ", omega))
  }

omegaSquared(1205.1, 774.4, 28.7, 2)
contrast1_effectSize = rcontrast(-5.978, 27)
contrast2_effectSize = rcontrast(-2.505, 27)
```

By calculating **omega squared**, $\omega^2$ and $r_{contrast}$,

$\omega^2 = \frac{SS_M - (df_M)MS_R}{SS_T + MS_R}$ , $r_{contrast} = \sqrt{\frac{t^2}{t^2+df}}$

we can report the result like:

-   All significant values are reported at p \< .05.There was a significant effect of teaching style on exam marks, F(2, 27) = 21.01, ω2 = .57. Planned contrasts revealed that reward produced significantly better exam grades than punishment and indifference, t(27) = --5.98, r = .75, and that punishment produced significantly worse exam marks than indifference, t(27) = −2.51, r = .43.

## SMT2

```{r}
superheroData <- read.delim("Superhero.dat", header = TRUE)
head(superheroData)
superheroData$hero <- factor(
  superheroData$hero, levels = c(1:4),
  labels = c("Spiderman", "Superman", "Hulk", "Teenage Mutant Ninja")
  )

head(superheroData)

superheroDatascatter <- ggplot(superheroData, aes(hero, injury))
superheroDatascatter +
  geom_point() +
  stat_summary(
    fun = mean, geom = "bar", fill = "white", color = "Black"
    ) +
  stat_summary(
    fun.data = mean_cl_normal, geom = "errorbar",
    width = 0.2, size = 0.75, color = "black"
    ) +
  stat_summary(fun = mean, geom = "point", size = 1, color = "black") +
  stat_summary(fun = mean, geom = "point", size = 1, color = "red") +
  labs(x = "Hero", y = "Injury")

leveneTest(superheroData$injury, superheroData$hero, center = median)

superheroModel <- aov(injury ~ hero, data = superheroData)
summary(superheroModel)
summary.lm(superheroModel)
```

The output of ANOVA indicates that there is a significant effect of superhero costume on injury severity.

However, we don't know the exact effect of superhero costume on injury severity. We don't know which group differed.

```{r}
pairwise.t.test(
  superheroData$injury, superheroData$hero, p.adjust.method = "bonferroni"
  )

omegaSquared <- function(ssm, ssr, msr, dfm)
  {
  omega <- (ssm - dfm*msr)/(ssm + ssr + msr)
  print(paste("omega^2 = ", omega))
  }

omegaSquared(4181, 4357, 167.6, 3)
```

Result

-   There was a significant effect of superhero costume on severity of injury, F(3, 26) = 8.32, $\omega^2 = .42$

-   Bonferroni test revealed that wearing a Superman costume resulted in significantly more injury than wearing a Hulk or Ninja turtle costume (p \< .05).

-   Wearing Spiderman or Hulk costume didn't resulted in significantly more injury than wearing any other costumes.

## SMT 4

Links between mobile phone use and brain tumours.

Factor: Six groups (0, 1, 2, 3, 4, 5 hours per day)

Outcome: Size

```{r}
tumourData <- read.delim("Tumour.dat", header = TRUE)
head(tumourData)
tumourData$usage <- factor(
  tumourData$usage, levels = c(0:5),
  labels = c("0H", "1H", "2H", "3H", "4H", "5H")
  )

tumourDatagraph <- ggplot(tumourData, aes(usage, tumor))
tumourDatagraph +
  geom_point() +
  stat_summary(
    fun = mean, geom = "bar", fill = "white", color = "Black"
    ) +
  stat_summary(
    fun.data = mean_cl_normal, geom = "errorbar",
    width = 0.2, size = 0.75, color = "black"
    ) +
  stat_summary(fun = mean, geom = "point", size = 1, color = "black") +
  stat_summary(fun = mean, geom = "point", size = 1, color = "red") +
  labs(x = "Usage", y = "Tumor")

leveneTest(tumourData$tumor, tumourData$usage, center = "median")

```

The assumption of homogeneity of variance has violated (p\<.05). Therefore, we have to transform the data or choose Welch's *F.*

```{r}
tumourModel <- aov(tumor ~ usage, data = tumourData)
summary(tumourModel)
```

The main ANOVA summary shows that there is a significant effect of mobile phone usage on the size of brain tumors. However, we still don't know the exact effect (which group differed).

`oneway.test()` doesn't assume the variances to be equal. (Still, we can set that the variances are equal by `var.equal = TRUE`)

```{r}
oneway.test(tumor ~ usage, data = tumourData, var.equal = FALSE)
```

The output shows Wehch's F, which is significant (p \< .05)

```{r}
pairwise.t.test(tumourData$tumor, tumourData$usage, p.adjust.method = "BH")

omegaSquared <- function(ssm, ssr, msr, dfm)
  {
  omega <- (ssm - dfm*msr)/(ssm + ssr + msr)
  print(paste("omega^2 = ", omega))
  }

omegaSquared(450.7, 38.1, 0.33, 5)
```

## SMT 5

One-way ANOVA to see if the change in hygiene is significantly different across people with different musical tastes.

Predictor: music

Outcome: change

```{r}
festivalData <- read.delim("GlastonburyFestivalRegression.dat", header = TRUE)

leveneTest(festivalData$change, festivalData$music, center = "median")

festivalModel <- aov(change ~ music, data = festivalData)
summary(festivalModel)

festivalData$music <- factor(
  festivalData$music, levels = c(1:4),
  labels = c("Crusty", "Indie Kid", "Metaller", "No Musical Affiliation")
  )

# Each category is compared to the last category, which is 'No Affiliation'
contrasts(festivalData$music) <- contr.SAS(4)
festivalContrast <- aov(change ~ music, data = festivalData, na.rm = TRUE)
summary.lm(festivalContrast)
```
