---
title: "SMT_Ch15"
author: "Daehyeog Lee"
format: html
editor: visual
---

```{r}
library(car); library(compute.es); library(effects); library(ggplot2)
library(multcomp); library(pastecs); library(WRS2); library(reshape)
library(here); library(ez); library(nlme); library(clinfun); library(pgirmess)
```

## SAT1

```{r}
MenDogs <- read.delim(here("ch15", "MenLikeDogs.dat"), header = TRUE)
head(MenDogs)

by(MenDogs[,2], MenDogs$species, stat.desc, basic=FALSE, norm=TRUE)

MenDogsModel<-wilcox.test(behaviou ~ species, data = MenDogs)
MenDogsModel

rFromWilcox(MenDogsModel, 40)
```

Men's and dogs' number of dog-like behaviors was not significantly different, $W = 205.5, p = 0.891$.

## SAT2

```{r}
darkLord <- read.delim(here("ch15", "DarkLord.dat"), header = TRUE)

stat.desc(darkLord, basic=FALSE, norm=TRUE)

darkModel <-
  wilcox.test(
    darkLord$message, darkLord$nomessag,  paired = TRUE, correct= FALSE
    )
darkModel

rFromWilcox(darkModel, 64)
```

The number of goats sacrificed after hearing the message was significantly less than after hearing the normal version of the song, $p = 0.04, r = -.26$.

## SAT3

```{r}
eastendersData <- read.delim(here("ch15", "Eastenders.dat"), header = TRUE)
stat.desc(eastendersData, basic = FALSE, norm = TRUE)

friedman.test(as.matrix(eastendersData))
friedmanmc(as.matrix(eastendersData))
```

-   The type of program watched significantly affected subsequent number of arguments, $\chi^2(2) = 7.59, p < .05.$

-   Watching EastEnders significantly affected the number of arguments compared to Friends (difference = 26.0). Other pairwise comparisons were not significant.

## SAT4

```{r}
clownData <- read.delim(here("ch15", "coulrophobia.dat"), header = TRUE)
head(clownData)

by(clownData$beliefs, clownData$infotype, stat.desc, basic=FALSE)


kruskal.test(beliefs ~ infotype, data = clownData)
clownData$Ranks <- rank(clownData$beliefs)
by(clownData$Ranks, clownData$infotype, mean)

kruskalmc(beliefs ~ infotype, data = clownData, cont = 'two-tailed')

jonckheere.test(clownData$beliefs, as.numeric(clownData$infotype))

#Relevel the variable infotype so that Adverts is level 1:
clownData$infotype<-factor(clownData$infotype, levels = levels(clownData$infotype)[c(2, 1, 3, 4)])

#Compare each group against Adverts:
kruskalmc(beliefs ~ infotype, data = clownData, cont = 'two-tailed')
```

-   Children's fear beliefs about clowns were significantly affected the format of information given to them, $H(3) = 17.06, p < .01$.

-   Paired comparisons of the mean ranks between groups showed that fear belief counts were significantly lower after exposure and stories, but not the control compared to watching adverts. A second set of comparisons revealed that compared to the control condition, fear belief scores were not significantly different after adverts, stories or exposure.
