---
title: "SMT_Ch9"
author: "Daehyeog Lee"
format: html
editor: visual
---

## Import Libraries

```{r}
library(ggplot2); library(pastecs); library(WRS2)
library(tidyr); library(dplyr); library(boot)
```

## Labcoat Leni's Real Research

```{r}
madData <- read.delim("Board&Fritzon(2005).dat", header = TRUE)
head(madData)

ttestfromMeans<-function(x1, x2, sd1, sd2, n1, n2)
{
  df <- n1 + n2 - 2
  poolvar <- (((n1 - 1) * sd1^2) + ((n2 - 1) * sd2^2)) / df
  t <- (x1 - x2) / sqrt(poolvar * ((1 / n1) + (1 / n2)))
  sig <- 2*(1 - (pt(abs(t), df)))
  paste("t(df = ", df, ") = ", t, ", p = ", sig, sep = "")
}

ttestfromMeans(
  madData$x1, madData$x2, madData$sd1, madData$sd2, madData$n1, madData$n2
  )
```

## SMT 1

```{r}
happyData <- read.delim("penis.dat", header = TRUE)
head(happyData)

happyData$book <- factor(
  happyData$book,
  levels = c(1:2),
  labels = c("Women are from Bras, Men are from Penis", "Marie Claire")
  )

happy.t.test <- t.test(happy ~ book, data = happyData)
happy.t.test

t <- happy.t.test$statistic[[1]]
df <- happy.t.test$parameter[[1]]
r <- sqrt(t^2/(t^2+df))
round(r, 3)
```

The happiness with people's relationship after reading "Marie Claire" is significantly higher than reading "Women are from..". The effect size of 0.451 is large enough.

## SMT 2

```{r}
happinessData <- read.delim("Field&Hole.dat", header = TRUE)
head(happinessData)

happiness.t.test <- t.test(
  happinessData$women, happinessData$statbook, paired = TRUE
  )
happiness.t.test

t <- happiness.t.test$statistic[[1]]
df <- happiness.t.test$parameter[[1]]
r <- sqrt(t^2/(t^2+df))
round(r, 3)
```

The result shows that reading "Women are from..." makes the readers happier than reading statistics textbook. The effect is statistically significant, but the size of the effect is small (0.12)
