---
title: "SMT_Ch15"
author: "Daehyeog Lee"
format: html
editor: visual
---

```{r}
library(car); library(compute.es); library(effects); library(ggplot2)
library(multcomp); library(pastecs); library(WRS2); library(reshape)
library(here); library(ez); library(nlme); library(clinfun); library(pgirmess)
library(MASS); library(mvoutlier); library(mvnormtest)
```

## Labcoat Leni

```{r}
marazillierData <- read.delim(here("ch16", "Marzillier & Davey (2005).dat"), header = TRUE)
head(marazillierData)

marazillierData$Induction <-
  factor(
    marazillierData$Induction,
    levels = c("Vignettes + Music", "Videos", "Memory Recall + Music")
    )
marazillierData$Mood <-
  factor(marazillierData$Mood, levels = c("Anxious", "Disgust", "Neutral"))

moodMelt <-
  melt(
    marazillierData,
    id = c("Induction", "Mood"),
    measured = c(
      "Anxiety.Change", "Sad.Change", "Happy.Change", "Angry.Change",
      "Disgust.Change", "Contempt.Change"
      )
    )
names(moodMelt) <- c("Induction", "Mood", "Outcome_Measure", "Frequency")

moodBar <- ggplot(moodMelt, aes(Mood, Frequency, fill = Outcome_Measure))
moodBar +
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") +
  stat_summary(
    fun.data = mean_cl_boot,
    geom = "errorbar",
    position = position_dodge(width=0.90), width = 0.2
    ) +
  labs(
    x = "Mood Induction Type",
    y = "Mean Change",
    fill = "Outcome Measure"
    ) +
  facet_wrap(~Induction, ncol = 3) +
  scale_y_continuous(breaks = seq(-40, 70, by = 10))

by(
  marazillierData$Anxiety.Change,
  list(marazillierData$Mood, marazillierData$Induction),
  stat.desc,
  basic = FALSE
  )
by(
  marazillierData$Sad.Change,
  list(marazillierData$Mood, marazillierData$Induction),
  stat.desc,
  basic = FALSE
  )
by(
  marazillierData$Happy.Change,
  list(marazillierData$Mood,marazillierData$Induction),
  stat.desc,
  basic = FALSE
  )
by(
  marazillierData$Angry.Change,
  list(marazillierData$Mood, marazillierData$Induction),
  stat.desc,
  basic = FALSE
  )
by(
  marazillierData$Disgust.Change,
  list(marazillierData$Mood, marazillierData$Induction),
  stat.desc,
  basic = FALSE
  )
by(marazillierData$Contempt.Change,
   list(marazillierData$Mood, marazillierData$Induction),
   stat.desc,
   basic = FALSE
   )


Mood_vs_None <- c(1, 1, -2)
anx_vs_disg  <- c(1, -1, 0)
contrasts(marazillierData$Mood) <- cbind(Mood_vs_None, anx_vs_disg)

Music_vs_None <- c(1, -2, 1)
vig_vs_mem <- c(1, 0, -1)
contrasts(marazillierData$Induction)<-cbind(Music_vs_None, vig_vs_mem)


outcome <-
  cbind(
    marazillierData$Anxiety.Change,
    marazillierData$Sad.Change,
    marazillierData$Happy.Change,
    marazillierData$Angry.Change,
    marazillierData$Disgust.Change,
    marazillierData$Contempt.Change
    )
marzillierModel <- manova(outcome ~ Induction*Mood, data = marazillierData)
summary(marzillierModel, intercept = TRUE)

summary.aov(marzillierModel)
anxiousModel <- lm(Anxiety.Change ~ Mood, data = marazillierData)
sadModel <- lm(Sad.Change ~ Mood, data = marazillierData)
happyModel <- lm(Happy.Change ~ Mood, data = marazillierData)
angryModel <- lm(Angry.Change ~ Mood, data = marazillierData)
disgustModel <- lm(Disgust.Change ~ Mood, data = marazillierData)
contemptModel <- lm(Contempt.Change ~ Mood, data = marazillierData)
summary.lm(anxiousModel)
summary.lm(sadModel)
summary.lm(happyModel)
summary.lm(angryModel)
summary.lm(disgustModel)
summary.lm(contemptModel)

anxiousModel2 <- lm(Anxiety.Change ~ Induction, data = marazillierData)
sadModel2 <- lm(Sad.Change ~ Induction, data = marazillierData)
happyModel2 <- lm(Happy.Change ~ Induction, data = marazillierData)
angryModel2 <- lm(Angry.Change ~ Induction, data = marazillierData)
disgustModel2 <- lm(Disgust.Change ~ Induction, data = marazillierData)
contemptModel2 <- lm(Contempt.Change ~ Induction, data = marazillierData)
summary.lm(anxiousModel2)
summary.lm(sadModel2)
summary.lm(happyModel2)
summary.lm(angryModel2)
summary.lm(disgustModel2)
summary.lm(contemptModel2)


moodBar <- ggplot(moodMelt, aes(Mood, Frequency, fill = Outcome_Measure))
moodBar +
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") +
  stat_summary(
    fun.data = mean_cl_boot,
    geom = "errorbar",
    position=position_dodge(width=0.90),
    width = 0.2
    ) +
  labs(
    x = "Mood Induction Type",
    y = "Mean Change",
    fill = "Outcome Measure"
    ) +
  scale_y_continuous(breaks = seq(-40, 70, by = 10))
```

Main effect of mood $F(12, 334) = 21.91, p < .001$, no significant main effect of the type of mood induction, $F(12, 334) = 1.12$. Not significant interaction $F(24, 676) = 1.12$

Follow-up univariate analysis

From the graph, changes in anxiety are higher over the three mood conditions. For disgust, the change is biggest after the disgust induction.

## SAT1

```{r}
chickenData <- read.delim(here("ch16", "chicken.dat"), header = TRUE)
head(chickenData)

chickenData$group <-
  factor(
    chickenData$group,
    levels = c(1:2),
    labels = c("Manic Psychosis", "Sussex Lecturers")
    )

chickenMelt <-
  melt(chickenData, id = c("group"), measured = c("quality", "quantity"))
names(chickenMelt) <- c("group","Outcome_Measure", "Value")

Boxplot <- ggplot(chickenMelt, aes(group, Value))
Boxplot +
  geom_boxplot() +
  labs(x = "Group", y = "Value") +
  facet_wrap(~ Outcome_Measure)

Bar <- ggplot(chickenMelt, aes(group, Value))
Bar +
  stat_summary(
    fun.y = mean, geom = "bar", position = "dodge", fill = "white"
    ) +
  stat_summary(
    fun.data = mean_cl_boot,
    geom = "errorbar",
    position = position_dodge(width=0.90),
    width = 0.2) +
  labs(x = "Group", y = "Value") +
  facet_wrap(~ Outcome_Measure, ncol = 2, scales = "free_y")

by(chickenData$quality, chickenData$group, stat.desc, basic = FALSE)
by(chickenData$quantity, chickenData$group, stat.desc, basic = FALSE)
by(chickenData[, 2:3], chickenData$group, cov)

Manic_Psychosis <- t(chickenData[1:10, 2:3])
Sussex_Lecturers <- t(chickenData[11:20, 2:3])

mshapiro.test(Manic_Psychosis)
mshapiro.test(Sussex_Lecturers)

aq.plot(chickenData[, 2:3])

contrasts(chickenData$group) <- c(-1, 1)

outcome <- cbind(chickenData$quality, chickenData$quantity)
chickenModel <- manova(outcome ~ group, data = chickenData)
summary(chickenModel, intercept = TRUE)
summary.aov(chickenModel)

qualityModel <- lm(quality ~ group, data = chickenData)
quantityModel <- lm(quantity ~ group, data = chickenData)

summary.lm(qualityModel)
summary.lm(quantityModel)


chickenDFA <- lda(group ~ quality + quantity, data = chickenData)
chickenDFA
plot(chickenDFA)
```

The main effect of `group` was significant (p = .032), which indicates that the groups differed in terms of the quality and quantity of their chicken impersonations.

However, the univariate analysis contradicted this result. The *p* values indicated that there were non-significant differences between groups in terms of both.

**From discriminant function analysis, both quality and quantity of impersonations have quite large coefficients, indicating that they both have a strong influence in discriminating the groups. However, they have the opposite sign, which suggests that group differences are explained by the difference between the quality and quantity of impersonations.**

## SAT2

```{r}

```

-   The type of program watched significantly affected subsequent number of arguments, $\chi^2(2) = 7.59, p < .05.$

-   Watching EastEnders significantly affected the number of arguments compared to Friends (difference = 26.0). Other pairwise comparisons were not significant.

## SAT4

```{r}

```

-   Children's fear beliefs about clowns were significantly affected the format of information given to them, $H(3) = 17.06, p < .01$.

-   Paired comparisons of the mean ranks between groups showed that fear belief counts were significantly lower after exposure and stories, but not the control compared to watching adverts. A second set of comparisons revealed that compared to the control condition, fear belief scores were not significantly different after adverts, stories or exposure.
