---
title: "SMT_Ch19"
author: "Daehyeog Lee"
format: html
editor: visual
---

```{r}
library(car); library(compute.es); library(effects); library(ggplot2)
library(multcomp); library(pastecs); library(WRS2); library(reshape)
library(here); library(ez); library(nlme); library(clinfun); library(pgirmess)
library(MASS); library(mvoutlier); library(mvnormtest)
library(corpcor); library(GPArotation); library(psych); library(DSUR.noof)
```

## Labcoat Leni

```{r}
dancerData = read.delim(here("ch19", "Miller et al. (2007).dat"),  header = TRUE)
head(dancerData)
Cyclephase_Factor <-
  factor(
    recode(dancerData$Cyclephase, "2=0;0=2"),
    levels = 0:2,
    labels = c("Fertile", "Menstural", "Luteal")
    )

Contraceptive_Factor <-
  factor(
    recode(dancerData$Contraceptive, "0=1;1=0"),
    levels = 0:1,
    labels = c("In Natural Cycle", "Using Pill")
    )

intercept <- gls(
  Tips~1, data = dancerData, method = "ML", na.action = na.exclude
  )
summary(intercept)


randomInt <-
  lme(
    Tips ~ 1,
    random = ~ 1|ID,
    data = dancerData,
    method = "ML",
    na.action= na.exclude
    )
summary(randomInt)
anova(intercept, randomInt)

cycleModel <- update(randomInt, .~. + Cyclephase_Factor, method = "REML")
pillModel <- update(cycleModel, .~. + Contraceptive_Factor)
finalModel <- update(pillModel, .~. + Cyclephase_Factor:Contraceptive_Factor)
anova(finalModel)
summary(finalModel)
intervals(finalModel)
```

## SAT1

```{r}
surgeryData = read.delim(here("ch19", "Cosmetic Surgery.dat"),  header = TRUE)

intercept <- gls(Post_QoL ~ 1, data = surgeryData, method = "ML")
randomIntercept <-
  lme(Post_QoL ~ 1, data = surgeryData, random = ~ 1 | Clinic, method = "ML")

randomInterceptSurgery <- update(randomIntercept, .~. + Surgery)
randomInterceptSurgeryQoL <- update(randomInterceptSurgery, .~. + Base_QoL)
addRandomSlope <- update(randomInterceptSurgeryQoL, random = ~ Surgery|Clinic)
addReason <- update(addRandomSlope, .~. + Reason)
finalModel <- update(addReason, .~. + Reason:Surgery)

BDIModel <- update(finalModel, .~. + BDI)
AgeModel <- update(BDIModel, .~. + Age)
genderModel <- update(AgeModel, .~. + Gender)

summary(genderModel)
anova(finalModel, genderModel)
```

`Age` significantly predicted quality of life after surgery but `Gender` didn't.

## SAT2

```{r}
satisfactionData = read.delim(
  here("ch19", "Honeymoon Period.dat"),  header = TRUE
  )
restructuredData <-
  reshape(
    satisfactionData,
    idvar = c("Person", "Gender"),
    varying = c("Satisfaction_Base", "Satisfaction_6_Months",
                "Satisfaction_12_Months", "Satisfaction_18_Months"),
    v.names = "Life_Satisfaction",
    timevar = "Time",
    times = c(0:3), 
    direction = "long"
    )

intercept <-
  gls(
    Life_Satisfaction ~ 1,
    data = restructuredData,
    method = "ML",
    na.action = na.exclude
    )
randomIntercept <-
  lme(
    Life_Satisfaction ~ 1,
    data = restructuredData,
    random = ~ 1 | Person,
    method = "ML",
    na.action = na.exclude,
    control = list(opt = "optim")
    )

timeRI <- update(randomIntercept, .~. + Time)
timeRS <- update(timeRI, random = ~Time|Person)
ARModel <- update(timeRS, correlation = corAR1(0, form = ~ Time|Person))
timeQuadratic <- update(ARModel, .~. + I(Time^2))
timeCubic <- update(timeQuadratic, .~. + I(Time^3))
genderModel <- update(timeCubic, .~. + Gender)
anova(timeCubic, genderModel)
summary(genderModel)
intervals(genderModel)
```

Including `Gender` as an additional covariate doesn't significantly improve the model.

## SAT3, 4

```{r}
#Task 3
exerciseData = read.delim(here("ch19", "Hill et al. (2007).dat"),  header = TRUE)

intercept <-
  gls(
    Post_Exercise ~ 1,
    data = exerciseData,
    method = "ML",
    na.action = na.exclude
    )
randomInt <-
  lme(
    Post_Exercise~1, data = exerciseData, random = ~ 1 | Classroom, method = "ML"
    )
intervention <- update(randomInt, .~. + Intervention)
anova(intercept, randomInt, intervention)
anova(intervention)
summary(intervention)

#Task 4
finalIntervention <- update(intervention, .~. + Pre_Exercise)
anova(intervention, finalIntervention)
summary(finalIntervention)
```

Adding the random effect of intercepts across classrooms results in a significantly better-fitting model. However, adding in the effect of intervention does not.

(Task 4) Adding the `Pre_Exercise` significantly improves the fit of the model.
